DSCI 445 - Algal Blooms

### 2020-12-10 paper.Rmd
Added subsections, figures, and captions to paper.

@@

## Methods TODO:
How we got the data

## Variable Selection

## Data Manipulation

@@

However, the aforementioned Interpolating approach certainly fails after a
certain range of days. The optimal shift was determined experimentally: Each
model was fit using data shifted from one to ten days in either direction, and
the cross-validation error was used to assess model accuracy. The specifics of
the experimental structure are described below. The model accuracy provided a
picture of the quality of interpolation provided by each shift. If data was
shifted too many days, the prediction accuracy of the algorithm would
deteriorate. FIXME mention optimal shift and include graph.

@@

The algae bloom data is spatiotemporal; the blooms are recorded in a certain
geographic location at a certain time. This makes the class of linear models
unsuitable for this problem. Linear models make assumptions about the structure
of the errors in the data generating process that are violated by spatiotemporal
data. Spatiotemporal data has errors which are correlated; as observations
closer in space or time will covary(CITE).

@@

>>>>

```{python, include=FALSE}
from plots import *
```

@@

TODO:
How we got the data

## Data Interpolation (Merging Shift Threshold)

@@

However, the aforementioned Interpolating approach certainly fails after a
certain range of days. The optimal shift was determined experimentally: Each
model was fit using data shifted from one to ten days in either direction, and
the cross-validation error was used to assess model accuracy. The specifics of
the experimental structure are described below. The model accuracy provided a
picture of the quality of interpolation provided by each shift. If data was
shifted too many days, the prediction accuracy of the algorithm should
deteriorate.

## Response Variable Transformation

As can be seen from Figure \ref{fig:resp-hist}, the chlorophyll variable, the response
variable in this case, was initially highly skewed. This had an impact on
initial prediction accuracy. In order to correct for the skewed nature of the
data, the response was log-transformed, yielding the second distribution in
Figure \ref{fig:resp-hist}.

```{python, echo=FALSE, fig.cap="Distribution of chlorophyll before and after log transformation.\\label{fig:resp-hist}"}
plot_histograms()
```

It should be noted that all MSE values were calculated from chlorophyll on the
log-transformed scale, then exponentiated to account for the variable
transformation.

@@

The algae bloom data is spatiotemporal; the blooms are recorded in a certain
geographic location at a certain time. This makes the class of linear models
unsuitable for this problem. Linear models make assumptions about the structure
of the errors in the data generating process that are violated by spatiotemporal
data. Spatiotemporal data has errors which are correlated; as observations
closer in space or time will covary (CITE).

@@

## Merge Shifting Threshold

Comparing across different shifting thresholds, all three models included in
the experiment performed similarly.

```{python, echo=FALSE, fig.cap="Cross-validation error for all three models at merge thresholds from 1 to 10.\\label{fig:thresh-performance}"}
plot_thresh_performance()
```

As we can see from Figure \ref{fig:thresh-performance}, the trend across merge
thresholds for each model is very similar. As can also be seen in Figure
\ref{fig:thresh-performance}, the optimal shifting threshold is three
days \ref{fig:thresh-performance}. It is also interesting to note that a shift of one day performed
worse than no shift, or a shift of two days. This is a counterintuitive result,
but it could be due to the tradeoff between the extra volume of data added by
larger shift thresholds versus the accuracy of the data at lower shifting
thresholds.

## Model Performance

All three models mentioned performed similarly, with no model greatly exceeding
the other two in terms of CV error. The best overall model, in terms of FIXME
error, was a random forest with a merge shift threshold of three days, that
included all features. The best model had a CV MSE of approximately 1.75, and a
test MSE of approximately 1.09.
